Assessing glymphatic transport velocities by adjoint methods

### 2018-10-09 tex/paper.tex
revised intro

\author{Lars Magnus Valnes, Sebastian K. Mitusch, Geir A. Ringstad, Per Kristian Eide, Simon W. Funke, Kent-Andre Mardal }

@@

A novel pathway called the paravascular pathway was discoved in 2011~\cite{iliff2012paravascular} and later found to be particularly active during sleep~\cite{xie2013sleep}.  
It has been proposed that this pathway plays an important role in the waste
clearance of the brain. In particular, since the brain lack a lymphatic system, 
which elsewhere in our body plays a crucial role in waste clearance, the system has been called the glymphatic system~\cite{jessen2015glymphatic}.  
The glymphatic system remains controversial and modeling attempts
has so far mostly failed to explain the biomechanics of the system~\cite{holter2017interstitial, smith2017glymphatic}. A proper biomechanical 
understanding of the pathway has significant potential because
dementia such as Alzheimer´s and Parkinson´s diseases are
associated with accumulation of metabolic waste such as
amyloid-beta and CSF-tau.  
In \cite{ringstad2018brain} brainwide distribution of MRI-contrast was demonstrated during 24 hours after lumbar contrast injection. Brain wide distribution by diffusion alone was deemed unlikely by the authors (some of which are part 
of this study as well). The argument was based on analytical considerations where it was calculated that 50\% contrast enrichment would occur after 
55 hours using the error function which is valid for planar diffusion. However, 
the surface of the brain is folded and is around five times larger than 
a corresponding surface of a ball with the same volume. Hence, 
a more rigorous modeling attempt is warrented. Complicating factors
are however that the contrast in the surrounding CSF is heterogenous
and changes significantly during the 24 hours of the investigations. 
Furthermore, images were obtained only at a few time-points during the investigations in \cite{ringstad2018brain}.  
Our purpose in this paper is to attempt a more rigorous methodology 
for assessing the apparent diffusion process. To this end we employ
adjoint methods to determine the apparent diffusion coefficient 
that would explain the tracer distribution.

@@

\subsection*{Mathematical Model}
In order to assess whether the contrast distribution shown 
in Figure \ref{fig1} is governed by a diffusion process we
aim to solve a PDE constrained optimization problem where
the contrast distribution, boundary conditions, and apparent
diffusion coefficient are solved for by optimizing with
respect to the observed contrast distribution. 
The objective function was defined as 
\begin{equation}
\min_{u,g} F = \quad \sum\limits_i\sp{n} \int\limits_{\Omega} |u(t_i) - u_{obs}(t_i)| \mathrm{d}\Omega + \frac{\alpha}{2} \int\limits_{0}\sp{T} || g ||_{L\sp{2}(\partial\Omega)} \mathrm{d}t + \frac{\beta}{2} \int\limits_{0}\sp{T} || \dot{g} ||_{L\sp{2}(\partial\Omega)}\mathrm{d}t 
\label{EQ::objf}
\end{equation}
subject to   
\begin{equation}
\begin{aligned}
\frac{\partial u}{\partial t} = \nabla \cdot  D_i \nabla u \qquad \text{in} \qquad \Omega \times \left\lbrace 0 , T \right)  \\
u=g(t) \qquad \text{on} \qquad \partial\Omega  \times \left\lbrace 0 , T \right) 
\end{aligned}
\label{Eq::PDE}
\end{equation}
with the domain $\Omega$ contains three sub domains, each with a different diffusion coefficient. We denote the Cerebral Spinal fluid (CSF) domain as $\Omega_1$, the grey matter as $\Omega_2$ and the white matter as $\Omega_3$.

@@

Noen detaljer her om element methode, optimerings algoritme, prekondisjonering
grid, osv.

>>>>

\author{Lars Magnus Valnes, Sebastian K. Mitusch, Geir A. Ringstad, \\ 
Per Kristian Eide, Simon W. Funke, Kent-Andre Mardal }

@@

A novel pathway called the paravascular pathway was discoved in 2011~\cite{iliff2012paravascular} and later found to be particularly active during sleep~\cite{xie2013sleep}.  
It has been proposed that this pathway plays an important role in the waste
clearance of the brain. In particular, since the brain lack a lymphatic system, 
which elsewhere in our body plays a crucial role in waste clearance, the system has been called the glymphatic system~\cite{jessen2015glymphatic}.  
The glymphatic system remains controversial and modeling attempts
has so far mostly failed to explain the biomechanics of the system~\cite{asgari2016glymphatic, holter2017interstitial, smith2017glymphatic}\kam{Lars: vi trenger flere her}. A proper biomechanical 
understanding of the pathway has significant potential because
dementia such as Alzheimer´s and Parkinson´s diseases are
associated with accumulation of metabolic waste such as
amyloid-beta and CSF-tau.

In \cite{ringstad2018brain} brainwide distribution of MRI-contrast was demonstrated during 24 hours after lumbar contrast injection. Brain wide distribution by diffusion alone was deemed unlikely by the authors (some of which are part 
of this study as well). The argument was based on analytical considerations where it was calculated that 50\% contrast enrichment would occur after 
55 hours using the error function which is valid for planar diffusion. However, 
the surface of the brain is folded and is around five times larger than 
a corresponding surface of a ball with the same volume. Hence, 
a more rigorous modeling attempt is warrented. Complicating factors
are however that the contrast in the surrounding CSF is heterogenous
and changes significantly during the 24 hours of the investigations. 
Furthermore, images were obtained only at a few time-points during the investigations in \cite{ringstad2018brain}.

The pioneering works of Sykov{\'a} and Nicholson~\cite{sykova2008diffusion} demonstrated
that diffusion was a governing transport mechanism in the brain and estimated to be on the order
of XXX. This is confirmed with DTI where values of young and healthy are typically around XXX
while values in dementia typically are XXX \kam{Geir vet du dette?}. Of course diffusion coefficient 
measured by DTI at a time-scale of XXX \kam{tall her} may not be representative for the process on a
longer time-scale, at least if the paravascular pathway plays an important role. The fact that the vasculature (mostly 
the capillary network) occupy around 3\% of the brain volume and the paravascular network is bound to be substatantially
smaller potentially render it invisible for the DTI investigation while still effective on longer time scales.

Our purpose in this paper is to attempt a more rigorous methodology 
for assessing the apparent diffusion process. To this end we employ
adjoint methods to determine the apparent diffusion coefficient 
that would explain the tracer distribution.

@@

\subsection*{Mathematical Model}
In order to estimate the apparent diffusion coefficient involved in the contrast transportation shown 
in Figure \ref{fig1} we assume that the process is governed by a diffusion equation and  
solve a PDE constrained optimization problem where
the contrast distribution, boundary conditions, and apparent
diffusion coefficient are solved for by optimizing with
respect to the observed contrast distribution. 
The objective function was defined as 
\begin{equation}
\min_{u,g} F = \quad \sum\limits_i\sp{n} \int\limits_{\Omega} |u(t_i) - u_{obs}(t_i)| \mathrm{d}\Omega + \frac{\alpha}{2} \int\limits_{0}\sp{T} || g ||_{L\sp{2}(\partial\Omega)} \mathrm{d}t + \frac{\beta}{2} \int\limits_{0}\sp{T} || \dot{g} ||_{L\sp{2}(\partial\Omega)}\mathrm{d}t 
\label{EQ::objf}
\end{equation}
subject to   
\begin{equation}
\begin{aligned}
\frac{\partial u}{\partial t} = \nabla \cdot  D_i \nabla u \qquad \text{in} \qquad \Omega \times \left\lbrace 0 , T \right)  \\
u=g(t) \qquad \text{on} \qquad \partial\Omega  \times \left\lbrace 0 , T \right) 
\end{aligned}
\label{Eq::PDE}
\end{equation}
with the domain $\Omega$ contains three sub domains, each with a different diffusion coefficient. We denote the Cerebral Spinal fluid (CSF) domain as $\Omega_1$, the grey matter as $\Omega_2$ and the white matter as $\Omega_3$.

@@

\kam{
Noen detaljer her om element methode, optimerings algoritme, prekondisjonering
grid, osv. 
}
